Reproducible Research: Peer Assessment 1
========================================================
## Loading and preprocessing the data

```{r load_data, echo=TRUE}
## Download .zip file
  fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, dest="activity.zip")
## unzip and load into memory
  fname <- unzip("activity.zip", list=FALSE)
  activity <- read.csv(fname, header=TRUE)
```
The variables in *activity* are :

**steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
**date**: The date on which the measurement was taken in YYYY-MM-DD format  
**interval**: Identifier for the 5-minute interval in which measurement was taken

## Mean and median total number of steps taken per day

Ignore the missing values in the dataset.
Histogram of the total number of steps taken each day
```{r histogram_Nsteps, fig.width=5, fig.height=4,echo=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 1, digits=1) #show 
steps <- aggregate(activity$steps, by=list(as.factor(activity$date)), FUN=sum)
colnames(steps) <- c("day","Nsteps")
hist(steps$Nsteps, main="Histogram of the number of steps each day",
     xlab="Number of steps", col="blue")
abline(v=median(steps$Nsteps, na.rm=TRUE), col="green", lwd=3)
meanSteps <- mean(steps$Nsteps, na.rm=TRUE)
medianSteps <- median(steps$Nsteps, na.rm=TRUE)
```
**Mean** total number of steps per day `r meanSteps`.  
**Median** total number of steps per day `r medianSteps`.  

## Average daily activity pattern

Create data frame *stepsInterval* with averaged number of steps for each interval.  

```{r daily_activity, fig.width=5, fig.height=4,echo=TRUE}
intervals <- list(as.factor(activity$interval))
stepsInterval <- aggregate(activity$steps, by=intervals, FUN=mean, na.rm=TRUE)
colnames(stepsInterval) <- c("Interval","Nsteps")
x <- as.numeric(stepsInterval$Interval)
plot(x, stepsInterval$Nsteps, type="l", lwd=2,
     main="Average number of steps per 5 min interval",
     xlab="Interval", ylab="Number of steps", col="blue")
maxInterval <- x[which.max(stepsInterval$Nsteps)]
```

The maximum number of steps, on average across all the days in the dataset, contains 5-minute interval number `r maxInterval`.

## Imputing missing variables

1. Copy *activity* frame to *activityImp*
2. Replace NAs in *activityImp* with the average number of steps for each interval from *stepsInterval*.

```{r count_NA}
# Make a logical vector *indexNA* of NAs
indexNA <- is.na(activity)
missingN <- sum(indexNA)
activityImp <-activity
for (i in 1:length(activityImp$steps)){
  if (indexNA[i]) 
         activityImp$steps[i]= stepsInterval$Nsteps[stepsInterval$Interval==activityImp$interval[i]]
}
```
Total of `r missingN` values were missing.
A histogram of the total number of steps taken each day with imputed data.
```{r histogram_Nsteps_Imputed, fig.width=5, fig.height=4,echo=TRUE}

stepsImp <- aggregate(activityImp$steps, by=list(as.factor(activityImp$date)), FUN=sum)
colnames(stepsImp) <- c("day","Nsteps")
hist(stepsImp$Nsteps, main="Histogram of the number of steps each day \n with imputed data",
     xlab="Number of steps", col="blue")
abline(v=median(stepsImp$Nsteps, na.rm=TRUE), col="green", lwd=3)
meanStepsImp <- mean(stepsImp$Nsteps)
medianStepsImp <- median(stepsImp$Nsteps)
```
**Mean Imputed** total number of steps per day `r meanStepsImp`.  
Same as **Mean** `r meanSteps`.  
**Median Imputed** total number of steps per day `r medianStepsImp`.  
Compare with **Median** `r medianSteps`.

## Activity patterns on weekdays and weekends.

Create a new factor column in *activityImp* with "day" levels.  
Rename the days as "weekday" and "weekend".
```{r factor_days}
activityImp$day <- as.factor(weekdays(as.Date(activityImp$date)))
activityImp$day <- sub("Monday|Tuesday|Wednesday|Thursday|Friday",
                       "weekday",activityImp$day)
activityImp$day <- sub("Saturday|Sunday", "weekend",activityImp$day)
```
Create dataframe *stepsDay* with averaged number of steps for each interval for the weekdays and weekends.  

```{r activity_days, fig.width=5, fig.height=4,echo=TRUE}
## Create dataframe stepsWeekday
intervalsWeekday <- list(as.factor(activityImp$interval[activityImp$day=="weekday"]))
stepsWeekday <- with (activityImp, aggregate(steps[day=="weekday"], 
                          by=intervalsWeekday,FUN=mean))
stepsWeekday$day <- factor("weekday")
intervalsWeekend <- list(as.factor(activityImp$interval[activityImp$day=="weekend"]))

## Create dataframe stepsWeekend
stepsWeekend <- with (activityImp, aggregate(steps[day=="weekend"], 
                          by=intervalsWeekend,FUN=mean))
stepsWeekend$day <- factor("weekend")

## Combine the two sets
stepsDay <- rbind(stepsWeekday, stepsWeekend)
colnames(stepsDay) <- c("Interval", "Nsteps", "day")
```
Activity on the weekdays and weekends
```{r activity_plot}
library(lattice)
x <- as.numeric(stepsDay$Interval)
with(stepsDay,xyplot(Nsteps ~ Interval| day, layout = c(1, 2), type="l"))
```

